Convergence rate of MAP estimates \\ for the exponential family

### 2020-11-27 main.tex
Change description of variance bound

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$. 
\begin{align}
	\Cov(g_t,\natp_{t+1}) 
	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \lr_t \sigma^2, \forall t \\
	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
\begin{align}
	g_t - \nabla f(\natp_t) = \E[T(X)] - T(X_{t+1})	
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.

>>>>

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps  
\begin{align}
	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. In our case, 
\begin{align}
	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.
